---
title: "MovieLens Report"
author: "Beauniah Kustaa"
date: "`r Sys.Date()`"
output: pdf_document
---

## Introduction

This project uses the MovieLens 10M dataset to build a recommendation system that predicts movie ratings using collaborative filtering methods. We use RMSE (Root Mean Squared Error) to evaluate our model's accuracy.

## Data Preparation

We begin by downloading and preparing the MovieLens dataset. We split it into `edx` for training and `final_holdout_test` for final evaluation, making sure there’s no leakage.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
```

```{r load-and-clean}
dl <- "ml-10M100K.zip"
if (!file.exists(dl)) {
  download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
}
ratings <- read_lines("ml-10M100K/ml-10M100K/ratings.dat")
movies <- read_lines("ml-10M100K/ml-10M100K/movies.dat")

ratings <- as.data.frame(str_split(ratings, fixed("::"), simplify = TRUE))
colnames(ratings) <- c("userId", "movieId", "rating", "timestamp")
ratings <- ratings %>%
  mutate(userId = as.integer(userId),
         movieId = as.integer(movieId),
         rating = as.numeric(rating),
         timestamp = as.integer(timestamp))

movies <- as.data.frame(str_split(movies, fixed("::"), simplify = TRUE))
colnames(movies) <- c("movieId", "title", "genres")
movies <- movies %>% mutate(movieId = as.integer(movieId))

movielens <- left_join(ratings, movies, by = "movieId")

set.seed(1, sample.kind = "Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

final_holdout_test <- temp %>%
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

removed <- anti_join(temp, final_holdout_test)
edx <- rbind(edx, removed)
```

## Data Exploration

Here we check the structure of the dataset — including number of users, movies, and how ratings are distributed.

```{r explore}
dim(edx)
n_distinct(edx$userId)
n_distinct(edx$movieId)
edx %>% count(rating) %>% arrange(desc(n))
```

## Model Development

We build our model in steps:
- Start with the global average rating
- Add **movie effects**
- Add **user effects**

```{r model-dev}
mu <- mean(edx$rating)

movie_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - mu))

user_avgs <- edx %>%
  left_join(movie_avgs, by = "movieId") %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))
```

## Model Evaluation on edx

We use RMSE to evaluate performance on the training dataset.

```{r eval-edx}
predicted_ratings <- edx %>%
  left_join(movie_avgs, by = "movieId") %>%
  left_join(user_avgs, by = "userId") %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)

rmse <- sqrt(mean((predicted_ratings - edx$rating)^2))
rmse
```

## Final Model and Holdout Evaluation

We apply the trained model to the `final_holdout_test` set and calculate the RMSE.

```{r eval-final}
final_predictions <- final_holdout_test %>%
  left_join(movie_avgs, by = "movieId") %>%
  left_join(user_avgs, by = "userId") %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)

final_rmse <- sqrt(mean((final_predictions - final_holdout_test$rating)^2))
final_rmse
```

## Conclusion

Our final model uses a simple yet effective approach combining global average, movie effects, and user effects. The RMSE achieved on the final holdout test set is:

**`r final_rmse`**

This model can be further improved by:
- Adding regularization to avoid overfitting
- Trying matrix factorization (`recosystem`)
- Using hybrid methods combining content and collaborative filtering
